---
title: "Mapping Lab"
date: "`r Sys.Date()`"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Unintentional Injuries Admissions

Today, we will be looking at real data from Public Health Scotland. You already know this dataset from last week's labs - it is called `unintentional_injuries.csv` and contains data from the 9 NHS Health Boards in Scotland about emergency hospital admissions as a result of unintentional injuries and assaults.

Our task today will be to plot this aggregated data in the form of a **map** of NHS boards, to gain better insights into unintentional injury admissions across Scotland. Aggregated data means that there are recorded observations of every combination of variable category.

### Task 1

Explore the Data Dictionary for this dataset here -\> <https://www.opendata.nhs.scot/dataset/unintentional-injuries/resource/aee43295-2a13-48f6-bf05-92769ca7c6cf>

### Task 2

**First** load the libraries that we need. (tidyverse, sf)

**Second** load the dataset and examine it. You may decide that you need to clean the names with `clean_names()` from the `janitor` package or you may decide they are ok. What would influence your decision?

```{r}
# Load the libraries we need

```

```{r}
# Load unintentional injuries data

```

### Task 3

**First** Filter the data to keep only the fall admissions where the patients are female, above 75 years old, and fell at home in 2022/23, for all HBR boards except "S92000003".

**Hint:** When you want to filter for rows that aren't equal to something, in R you can use the != operator. We want to remove the rows where HBR is "S92000003" because that is an aggregated row of all Health Boards.

**Second** Group your filtered admissions by Health Board and create a numerical summary to compute the sum of number of admissions. Save to an object called `falls_admissions_75_summary`

```{r}

```

### Task 4

**First** Load the NHS Health board Shapefile (NHS_healthboards_2019.shp) and save it into an object called `NHS_healthboards` using `st_read`. Examine the `NHS_healthboards` object. What is the geometry type?

**Second** Join the `NHS_healthboards` data with your filtered and grouped admissions data `falls_admissions_75_summary`.

```{r}

```

### Task 5

Plot a map of the number of falls at home in the female 75+ age group across NHS health boards in Scotland.

**Hint:** Use ggplot + geom_sf, and aesthetically fill with the number of falls. Try using scale_fill_distiller with a palette. Add a title and a subtitle. Use your plotting skills to make it look clear and understandable. E.G Try using some theme settings.

What do the colours mean? How would you interpret this map?

```{r}

```

### Task 6

The problem with using the raw number of falls is that the health boards have different sizes - the raw number does not tell us how prevalent falls are in the local population. So let's formulate a new question.

**Question: In a group of 10,000 women over 75, how many can we expect to be admitted to hospital following a fall at home in each health board?**

Using the same process as above, look at the number of falls per 10,000 women aged 75 and over for falls at home per health board and plot that to a map. 

Ensure your map is a different colour to the previous one. You can use previous objects that you created that filtered falls at home for women over 75 and the `falls_admissions_75_summary` in your solution.

You will need to read in the `hb2019_pop_est.csv` which is a dataset of population estimates for all 14 health boards (2019) in Scotland from 1981 to 2022. The data dictionary link is here:

<https://www.opendata.nhs.scot/dataset/population-estimates/resource/27a72cc8-d6d8-430c-8b4f-3109a9ceadb1>

**First** Read in the population data and filter for the year 2022, Female and remember to filter out the whole of Scotland code S92000003. Add a new column that shows the sum of all the columns across Age75 to Age90plus. Hint: `rowSums` is a handy function.

**Second** Join the population data with `falls_admissions_75_summary`, the object you created earlier

**Third** Create a new column that is the falls per 10k

**Fourth** Join with the NHS_healthboards data to get the GEOMETRY column and health board name

**Fifth** Plot a map using ggplot and geom_sf

```{r}

```

### Task 7

Maps should have a north arrow and a scale so we're going to add one to one of our maps. 

**First** install and load the `ggspatial` library and have a browse if how it works here -> https://paleolimbot.github.io/ggspatial/

**Second** choose one of your maps and add the following snippet of code to the plot. Spend a bit of time customising your arrow, scale and map to make it appear as you want it to. 

```{r}
  annotation_scale(    
    location = "tl"
  ) +
  annotation_north_arrow(
    location = "tl",    
    pad_y = unit(0.5, "in"),    
    style = north_arrow_nautical(
      fill = c("grey40", "white"),      
      line_col = "grey20"
    ) 
  )
```
