---
title: "DRY coding"
author: "Data Science in Biomedicine - Week 5"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
## Introduction

We'll be using an in-built data set from the MASS package. But, because the MASS package can be a bit problematic to install, I've provided the data for you on Learn. It's in .rds format, so we load it in using the `read_rds()` function.

For more info about the dataset, head to https://stat.ethz.ch/R-manual/R-devel/library/MASS/html/survey.html

```{r}
library(tidyverse)
library(knitr)

survey <- read_rds("./data/survey.rds")
```

## Exercise 1
Explore the data - what variables are in there, what variable types, how many variables and observations?
```{r}
glimpse(survey)
```

## Exercise 2
Round every numeric variable to 0 decimal places.
```{r}

survey_rounded <- survey %>% 
  mutate(across(where(is.numeric), ~round(.x, digits = 0)))
```


## Exercise 3
Using the rounded dataframe you created in Exercise 2, calculate the mean of each numeric variable in the dataset. Use kable() to insert a nicely formatted table into your R Markdown document.

```{r}
means <- survey_rounded %>% 
  select(where(is.numeric)) %>% 
  map_df(~mean(.x, na.rm = TRUE)) %>% 
  kable()

means

```

## Exercise 4
Now make exactly the same table, using the pivot functions. Again, use kable() to insert a nicely formatted table into your R Markdown document.

```{r}
means_pivot <- survey_rounded %>% 
  select(where(is.numeric)) %>% 
  pivot_longer(cols = everything()) %>% 
  group_by(name) %>% 
  summarise(mean = mean(value, na.rm = TRUE)) %>% 
  arrange(desc(name)) %>% 
  pivot_wider(names_from = name, values_from = mean) %>% 
  kable()

means_pivot
```

## Exercise 5
Students had initially expressed their height in either imperial or metric units, but someone converted everything into centimeters. Create a new variable, Height_original, which will show height in centimeters for students who expressed their height in metric units, and height in inches for students who prefer imperial units. 

NOTE: This one is a bit of a challenge. You may want to check out the `case_when()` function.

```{r}
survey <- survey %>% 
  mutate(Height_original = case_when(M.I == "Metric" ~ Height,
                                     M.I == "Imperial" ~ Height/2.54))

# Alternatively, to add the unit 
survey_with_unit <- survey %>% 
  mutate(Height_original = case_when(M.I == "Metric" ~ paste0(Height, " cm"),
                                     M.I == "Imperial" ~ paste0(round(Height/2.54), " in")))
```


## Exercise 6
Create a table that shows the mean and standard deviation of height - in centimeters for students who used metric units, and in inches for students who used imperial units. Use kable() to insert a nicely formatted table into your R Markdown document.

```{r}
survey %>% 
  filter(!is.na(M.I)) %>% 
  group_by(M.I) %>% 
  summarise(mean_height = mean(Height_original),
            sd_height = sd(Height_original)) %>% 
  kable()
```


## Exercise 7
Create a contingency table to see the relationship between exercise and smoking. Pivot your table to make it easier to read, and then use kable() to insert a nicely formatted table into your R Markdown document.

```{r}
survey %>% 
  group_by(Smoke, Exer) %>% 
  tally() %>% 
  pivot_wider(names_from = Exer, values_from = n) %>% 
  kable()
```


## Exercise 8
Create a grid of Pulse histograms - one histogram for each combination of Sex and Exercise.

```{r}

# The function we can use here is facet_grid()
survey %>% 
  filter(!is.na(Sex)) %>% 
  ggplot(aes(x = Pulse)) +
  geom_histogram() +
  facet_grid(Sex ~ Exer)

survey %>% 
  mutate(Exer = fct_relevel(Exer, "None", "Some", "Freq")) %>% 
  filter(!is.na(Sex)) %>% 
  ggplot(aes(x = Pulse)) +
  geom_histogram() +
  facet_grid(Sex~Exer)
```


## Exercise 9
Create a new variable, Hnd.diff, that shows the difference in span between the writing and non-writing hand. Create a set of boxplots of this variable, with a separate boxplot for those who write with their right hand, and another one for those who write with their left hand.

```{r}
# Add the new variable
survey <- survey %>% 
  mutate(Hnd.diff = Wr.Hnd - NW.Hnd)

# Create box plots (removing NAs first)
survey %>% 
  filter(!is.na(W.Hnd)) %>% 
  ggplot(aes(x = W.Hnd, y = Hnd.diff)) +
  geom_boxplot() +
  labs(title = "Difference in span between writing and non-writing hand",
       subtitle = "By dominant (writing) hand",
       x = "Writing hand",
       y = "Difference in span")
```

