---
title: "Unintentional Injuries"
author: "Data Science in Biomedicine - Week 6"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Background

Unintentional injury is one of the leading causes of death in children and young adults. The elderly are also more vulnerable, and it is a common cause of emergency hospital admission in people of all ages. Many injuries may be preventable, and through the collection and analysis of data, patterns can be explored, and evidence gathered, helping to inform policy makers and the public so that useful evidence-based prevention strategies can be developed.

In this project you will be exploring the **Unintentional Injuries** datasets (admissions and deaths) from the Public Health Scotland open data platform. The datasets include information on admissions and deaths relating to injuries and assaults in Scotland.

Here is a link with more information on the **Unintentional Injuries** datasets:

-   <https://www.opendata.nhs.scot/dataset/unintentional-injuries>

## Data

There are three data files that you need to download from the URLs are below. Click on the URL of the CSV files to download each file.

-   <https://www.opendata.nhs.scot/dataset/unintentional-injuries/resource/aee43295-2a13-48f6-bf05-92769ca7c6cf>
-   <https://www.opendata.nhs.scot/dataset/unintentional-injuries/resource/89807e07-fc5f-4b5e-a077-e4cf59491139>
-   <https://www.opendata.nhs.scot/dataset/geography-codes-and-labels/resource/652ff726-e676-4a20-abda-435b98dd7bdc>

Move the files into a data folder. You should have 3 files.

1.  ui_admissions_2023.csv
2.  ui_deaths-2023-.csv
3.  hb14_hb19.csv

A few points to note about the dataset:

-   Many of the rows contain aggregated data, due to each variable including categories such as “All”. You will need to decide what to do with this aggregate level data.
-   The variable “HBR” refers to Health Board Region codes, which is the variable to match if joining with the Health Boards look-up. Be sure to check for aggregate level codes in the Health Board look-up as well.
- The Health Board code for the whole of Scotland is S92000003.

## Your task

Your task today is to explore the injuries datasets and answer the questions below. To answer the questions you must use the datasets provided. The R code used in all aspects of loading, processing, analysis and reporting must be visible in your final submission. You are encouraged to include tables and figures to illustrate your report. You can use any R packages of your choosing to create this report; you are not restricted to those discussed in the course.

## The output

In your report, you must explore the following questions:

-   What types of injuries are most common and for which demographic (the data include age and sex)?
-   What is the rate of death in admissions of these injuries?

Your report should be written as if it were to be read and followed by another student on this course. Please explain your analytical decisions - it's helpful to know why you decided to do things in this way and not another. Add comments in your code, especially when you use functions or approaches not covered in this course.

We are not looking for one specific analysis. It is up to you to decide how to operationalise the two research questions. As you work with the data, feel free to make decisions on whether you want to focus on a specific demographic or location(s) if that makes your message clearer and analysis more sound.

## The MVP for question 1

You can go quite far towards answering the first question without doing any joins or complicated data wrangling, and using just the `ui_admissions_2023.csv` file. Explore the data file, deal with the aggregate data rows, and produce a visualisation that answers (one aspect of) this question.

NOTE: It's probably a good idea to focus on a single year at first. You'll have to make a decision about which year it should be - remember to justify your decision in writing.

```{r}
library(tidyverse) 
library(janitor) 
library(kableExtra)

admissions_data <- read_csv("./data/ui_admissions_2023.csv")

# Focusing on the most recent data (2022/23)
admissions_for_MVP <- admissions_data %>% 
  # Look at only the latest year
  filter(FinancialYear == "2022/23") %>%
  # Look at only the whole of Scotland
  filter(HBR == "S92000003") %>%
  # All injury locations
  filter(InjuryLocation == "All") %>%
  # Select justthe columns we want
  select(AgeGroup, Sex, InjuryType, NumberOfAdmissions) %>% 
  # Filter out the Sex and AgeGroup aggregated data
  filter(Sex != "All" & AgeGroup != "All") %>% 
  # Filter out the All Diganoses agreggated data
  filter(InjuryType != "All Diagnoses") %>%
  clean_names()

admissions_sex_bar_chart <- admissions_for_MVP %>%
  ggplot(aes(x = reorder(injury_type, number_of_admissions), 
              y = number_of_admissions, fill = sex)) +
  # We could use geom_col() here
  geom_bar(stat = "identity") + 
  # Flip the axes of the plot
  coord_flip()

admissions_sex_bar_chart
```

```{r}
admissions_age_bar_chart <- admissions_for_MVP %>%
  ggplot(aes(x = reorder(injury_type, number_of_admissions), 
              y = number_of_admissions, fill = age_group)) +
  geom_bar(stat = "identity", position = "dodge") + 
  coord_flip()

admissions_age_bar_chart

admissions_table_MVP <- admissions_for_MVP %>% 
  group_by(sex, age_group) %>%
  summarise(admissions = sum(number_of_admissions)) %>% 
  kable()

admissions_table_MVP
```

## One step further

Now that you've gotten to know the admissions data a little, think about another visualisation or table that would go further towards answering the first question. Don't create it now, just describe what you would want to do.

```{r}
admissions_sex_age_table <- admissions_for_MVP %>%
  # We want to reorder and rename the age_group
  mutate(age_group = factor(age_group)) %>%
  mutate(age_group = fct_relevel(age_group, "5-9 years", after = 1)) %>% 
  mutate(age_group = fct_recode(age_group, "75+ years" = "75plus years")) %>% 
  group_by(sex, age_group) %>%
  summarise(admissions = sum(number_of_admissions)) %>% 
  # Pivot wider for a nice table
  pivot_wider(names_from = "sex", values_from = "admissions") %>%
  kable(col.names = c("Age Group", "Female", "Male"), 
        align = "lcc",
        digits = 2,
        caption = "Unintentional injury admissions in 2022/23") %>%
  kable_styling("striped", full_width = FALSE) %>% 
  footnote(general = "Data from Public Health Scotland")

admissions_sex_age_table
```

## The join

In order to answer the second question, you will need to join the admissions file with the deaths file. Have a look at both files first, to identify the potential issues. Fix the issues, then go ahead and join.

```{r}
deaths_data <- read_csv("./data/ui_deaths_-2023-.csv")

# If you simply run a join without any preparation, the result will be a very long file
# with a warning about many-to-many relationships. This is a sign that something
# didn't go right.

# First issue: Year is recorded differently in the two files. The solution is to extract # the first part of the FinancialYear variable (up to the / character) in the admissions # data. There are various way to do this, one example below:
admissions_data_for_join <- admissions_data %>%
  separate(FinancialYear, into = c("Year", NA), sep = "/") %>% 
  mutate(Year = as.numeric(Year))

# Second issue: Injury types in the deaths data are slightly different to the ones in
# admissions data (but still similar enough that you can figure out how they match).
# Let's recode the names in the deaths data, so they match the names in the admissions
# data:
deaths_data_for_join <- deaths_data %>%
  mutate(InjuryType = factor(InjuryType)) %>%
  mutate(InjuryType = fct_recode(InjuryType, "RTA" = "Land transport accidents",
                                 "All Diagnoses" = "All",
                                 "Accidental Exposure" = "Accidental exposure",
                                 "Struck by, against" = "Struck by,against"))
# And now we can join:
joined_data <- full_join(admissions_data_for_join, deaths_data_for_join)
```

## The MVP for question 2

What would the MVP be for question 2? Create it. And then describe how you would take things further, if you had the time.

```{r}
# I will focus just on the aggregate data (all age groups, all sexes, all locations):
deaths_table <- joined_data %>%
  mutate(DeathRate = NumberofDeaths/NumberOfAdmissions) %>%

  # Note the filter below - I'm keeping only the aggregate rows 
  filter(AgeGroup == "All", Sex == "All", InjuryLocation == "All") %>% 
  filter(Year == 2022, HBR == "S92000003") %>%
  select(InjuryType, DeathRate) %>%
  kable(
    col.names = c("Injury Type", "Death Rate"),
    digits = 2,
    caption = "Death rates from unintentional injuries in Scotland, 2022"
  ) %>% 
  kable_styling("striped", full_width = F)

deaths_table
```


## Use all information you've got

The exercise authors provided you with `hb_lookup.csv`, which gives you the names of health boards. Join it with your main data file and make use of it. Here are some examples of what you could do:

-   focus on a single health board
-   compare and contrast two health boards
-   compare all health boards

```{r}
health_boards <- read_csv("./data/hb14_hb19.csv") %>% 
  select(HB, HBName)

joined_data <- joined_data %>% 
  full_join(health_boards, by = c("HBR" = "HB"))


deaths_table_lothian <- joined_data %>%
  filter(AgeGroup == "All", Sex == "All", InjuryLocation == "All") %>% 
  filter(Year == 2022, HBName == "NHS Lothian") %>%   
  group_by(InjuryType) %>%
  summarise(NoOfDeaths = sum(NumberofDeaths, na.rm = TRUE),
  NoOfAdmissions = sum(NumberOfAdmissions, na.rm = TRUE)) %>% 
  ungroup() %>%
  mutate(DeathRate = NoOfDeaths/NoOfAdmissions) %>%
  kable(
    col.names = c("Injury Type", "No. of Deaths", "No. of Admissions", "Death Rate"),
    digits = 2,
    caption = "Death rates from unintentional injuries, NHS Lothian 2022"
  )

deaths_table_lothian
```


## Knit

At this point, knit your report into PDF and see if it looks nice. If you spot any issues, fix them or ask for help now.

## Do more

If you have time, try and implement some of your "one step further" ideas from above.
