---
title: "Week 02 - dplyr - SOLUTIONS"
author: "Kasia Banas"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
```

## Exercise 01: select

Use [diamonds](https://ggplot2.tidyverse.org/reference/diamonds.html) example table from `ggplot2` package.

Select only `price`, `cut`, and `color` columns. Use piping and `select`.

```{r exercise 01-1}
data("diamonds")
diamonds %>% 
  select(price, cut, color)
```

Select only `price`, `cut`, and `color` columns. Use base R (no piping or select).

```{r exercise 01-2}
diamonds[,c("price", "cut", "color")]
```

Select all columns except for `x`, `y`, and `z`. Use negative selection and piping.

```{r exercise 01-3}
diamonds %>% 
  select(-c(x, y, z))
```

Check out the help page for the `select()` function. Use `:` notation to select columns from `carat` till `clarity`.

```{r exercise 01-4}
diamonds %>% 
  select(carat:clarity)
```

And now select these same columns by capitalising on the fact that all their names (and none of the others) start with the letter "c".

```{r exercise 01-5}
diamonds %>% 
  select(starts_with("c"))
```

BONUS: How would you select these 4 columns in base R, without simply listing their names?

```{r exercise 01-6}
# Select columns that start with "c"
selected_columns <- diamonds[, startsWith(names(diamonds), "c")]

# Print the first few rows of the selected columns
print(head(selected_columns))
```

## Exercise 02: conditions 1

Examine the code and decide on whether a condition will be `TRUE` or `FALSE`. Write down the answer and run the chunk to check the answer.

**Answer 1:**

```{r exercise 02-1}
x <- 4
(x > 3)
```

TRUE

**Answer 2:**

```{r exercise 02-2}
x <- 4
y <- 5
(x != y)
```

TRUE

## Exercise 03: conditions 2

Examine the code and decide on whether a condition will be `TRUE` or `FALSE`. Write down the answer and run the chunk to check the answer.

**Answer 1:**

```{r exercise 03-1}
x <- 4
!(x < 3)
```

TRUE

**Answer 2:**

```{r exercise 03-2}
x <- 4
y <- 5
!(x != y)
```

FALSE

## Exercise 04: conditions 3

Examine the code and decide on whether a condition will be `TRUE` or `FALSE`. Write down the answer and run the chunk to check the answer.

**Answer 1:**

```{r exercise 03-1}
x <- 4
(x < 3) | (x >=4)
```

TRUE

**Answer 2:**

```{r exercise 04-2}
x <- 4
y <- 6
(x == y) | ((x < 5) & (y > 5))
```

TRUE

**Answer 3:**

```{r exercise 04-3}
x <- 3
y <- 3
(x != y) & ((x <= 3) & (y >= 3))
```

FALSE

**Answer 4:**

```{r exercise 04-4}
x <- 3
y <- 3
(x != y) | ((x <= 3) & (y >= 3))
```

TRUE

## Exercise 05: conditions

Examine the code and decide on whether a condition will be `TRUE` or `FALSE`. Write down the answer and run the chunk to check the answer.

**Answer 1:**

```{r exercise 05-1}
x <- seq(-4, 4, 2)
(x ==0)
```

F F T F F

**Answer 2:**

```{r exercise 05-2}
x <- seq(-4, 4, 2)
y <- 1:5
(x + y <= 0)
```

T T F F F

**Answer 3:**

```{r exercise 05-3}
x <- seq(-4, 12, 4)
y <- 1:5
(x > y)
```

F F T T T

**Answer 4:**

```{r exercise 05-4}
x <- seq(-4, 12, 4)
y <- 1:5
(x > y) & ((x < 10) | (y < 3))
```

## Exercise 06: logical indexing

Use logical indexing to values of `x` that are above 3.

```{r exercise 06-1}
x <- -3:7
# Your code goes here
x[x>3]
```

Use logical indexing to extract values of `x` when it is higher than value in `y`

```{r exercise 06-2}
x <- -3:7
y <- seq(5, -2, length.out = 11)
# Your code goes here
x[x>y]
```

Use logical indexing to extract values of `x` that are greater than the mean value of x (avoid using intermediate variables).

```{r exercise 06-3}
x <- -3:7
# Your code goes here
x[x>mean(x)]
```

## Exercise 07: filter

Use [msleep](https://ggplot2.tidyverse.org/reference/msleep.html) example table from `ggplot2` package.

Filter the table to keep only `"domesticated"` animals (variable name is `conservation`).

```{r exercise 07-1}
data(msleep)
msleep %>% 
  filter(conservation == "domesticated")
```

Filter the table to keep only the omnivores (`vore`) that sleep more than 10 hours in total.

```{r exercise 07-2}
msleep %>% 
  filter(vore == "omni" & sleep_total >10)
```

Filter the table to exclude animals with body weight above `50`.

```{r exercise 07-3}
msleep %>% 
  filter(bodywt <= 50)
```

Filter the table to exclude animals with body weight below [median](https://stat.ethz.ch/R-manual/R-devel/library/stats/html/median.html) body weight. Hint, you do not need a temporary variable.

```{r exercise 07-4}
msleep %>% 
  filter(bodywt >= median(bodywt))
```

## Exercise 08: arrange

Use [presidential](https://ggplot2.tidyverse.org/reference/presidential.html) example table from *ggplot2* package.

Arrange the table by party.

```{r exercise 08-1}
data("presidential")
presidential %>% 
  arrange(party)
```

Arrange the table by start date in the descending order.

```{r exercise 08-2}
presidential %>% 
  arrange(desc(start))
```

Arrange the table by party (descending order) and president's name (ascending order).

```{r exercise 08-3}
presidential %>% 
  arrange(desc(party), name)
```

## Exercise 9: mutate

Use [msleep](https://ggplot2.tidyverse.org/reference/msleep.html) example table from `ggplot2` package.

Compute new variable / column with sleep to body-weight ratio.

```{r exercise 10-1}
msleep_with_ratio <- msleep %>% 
  mutate(sleep_wt_ratio = sleep_total/bodywt)
```

Compute new variable / column `is_big` of logical value that are `TRUE` if body_weight is above `50`.

```{r exercise 10-2}
msleep_with_big <- msleep %>% 
  mutate(is_big = bodywt > 50)
```

Compute three new variables / columns. First, compute the *mean* and *standard deviation* of total sleep variable. You will get the same value in all the rows, make sure that your columns are *not* called `mean` and `sd`. Then compute a z-score of the total sleep: $Z(x) = (x - mean(x)) / sd(x)$. Put all computations into a single mutate code.

```{r exercise 10-3}
msleep_with_sleep_z <- msleep %>% 
  mutate(mean_sleep = mean(sleep_total, na.rm = TRUE),
         sd_sleep = sd(sleep_total, na.rm = TRUE),
         z_sleep = (sleep_total - mean_sleep)/sd_sleep)
```

Compute z-score for body weight variable, do not store mean and standard deviation in extra columns but use them directly.

```{r exercise 10-4}
msleep_with_bodywt_z <- msleep %>% 
  mutate(z_bodywt = (bodywt - mean(bodywt, na.rm = T))/sd(bodywt, na.rm = T))
```
