---
title: "Week02_TBI"
author: "Christopher A Oldnall"
date: "`r Sys.Date()`"
output: html_document
editor_options: 
  chunk_output_type: inline
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

You will need to load the required libraries first: tidyverse and here.

```{r libraries}
# install.packages("tidyverse")
# install.packages("here")
library(tidyverse)
library(here)
```

We will be using traumatic brain injury data from a Tidy Tuesday challenge
Learn more about the challenge and the data here (spend max 3 minutes looking at the linked page):
https://github.com/rfordatascience/tidytuesday/blob/master/data/2020/2020-03-24/readme.md

Let's start by loading in the data

```{r loading-data}
tbi_age <- read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-03-24/tbi_age.csv')
tbi_year <- read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-03-24/tbi_year.csv')
```

Now, set up folders in your working directory: raw_data, scripts, plots and tables. Put this RMarkdown document in your scripts folder. 

### Task 1
Save both datasets in the raw_data folder (use the 'here' library)

```{r saving-data-locally}
write_csv(tbi_age, file= here("raw_data", "tbi_age.csv"))
write_csv(tbi_year, file= here("raw_data", "tbi_year.csv"))
```


### Task 2
Use your favourite functions for eyeballing data and write a couple of sentences about each dataset (its dimensions, the variables inside, types of data, etc.)

```{r Task 2}
str(tbi_age)
table(tbi_age$type)
table(tbi_age$injury_mechanism)

str(tbi_year)
table(tbi_year$injury_mechanism)
table(tbi_year$type)
```

tbi_age:
This data frame has 231 records and 5 variables. The variables are age_group (aggregated into categories currently stored as characters), type which represents what was the result of the brain injury (also split into 3 categories but currently strings), injury_mechanism (also split into 8 categories but currently strings). number_est is then the number within that specific age/type/mechanism group whilst the rate_est is the number_est divided by the total population. Both of these are continious numbers.

tbi_year:
This data has 216 records and 5 variables. There is an injury_mechanism variable (currently strings but with the same 8 categories), then type which is also analogous to the previous data frame. We then have year, rate_est and number_est which are all currently saved as continious numbers.

### Task 3
Imagine that you are preparing a brief for the local health authority, and they have asked you to answer the following questions. For each question, write a chunk of code that would provide the answer in a clean and reproducible way.

### Task 3.1
Which injury mechanism caused the most deaths in total?

```{r most-deaths}
tbi_age %>% 
  filter(type == "Deaths" & age_group == "Total") %>% 
  arrange(desc(number_est)) %>% 
  head(n = 1)
```

### Task 3.2
Looking just at the totals, what is the number of ED visits, hospitalisations and deaths
associated with each injury mechanism? Present the results as a table. Save this table in your tables folder.

```{r totals-table}
totals_table <- tbi_age %>%
  filter(age_group == "Total") %>% 
  select(-rate_est) %>% 
  pivot_wider(names_from = "type", values_from = "number_est") %>% 
  select(-age_group) %>% 
  arrange(injury_mechanism)

write_tsv(totals_table, file = here("tables", "totals_table.tsv"))
write_csv(totals_table, file = here("tables", "totals_table.csv"))
```
### Task 3.3
Focusing on the oldest age group (75+), draw a bar chart illustrating the number of hospitalisations associated with each injury mechanism. Make sure that you add a title and appropriate axis labels. Save this chart in your plots folder.

```{r oldest-hospitalisations}
tbi_age$injury_mechanism <- str_wrap(tbi_age$injury_mechanism, width = 15)

oldest_hosp_plot <- tbi_age %>% 
  filter(age_group == "75+" & type == "Hospitalizations") %>% 
  ggplot(aes(x = injury_mechanism, y = number_est)) +
  geom_col() +
  scale_x_discrete(guide = guide_axis(angle = 90)) +
  ggtitle("Plot showing hospitalisation numbers for each injury mechanism (Age Group: 75+)") +
  xlab("Injury Mechanism") +
  ylab("Estimated Number")

# Chris Soln (using png)
png(filename="../plots/oldest_hosp_plot.png")
oldest_hosp_plot
dev.off()

# Kasia Soln (using ggsave)
ggsave(filename = "../plots/oldest_hosp_plot.png", plot = oldest_hosp_plot)
```


