---
title: "Week 1 Lab 1"
author: "Insert your name here"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Getting started

The Behavioral Risk Factor Surveillance System (BRFSS) is an annual telephone 
survey of 350,000 people in the United States. As its name implies, the BRFSS 
is designed to identify risk factors in the adult population and report 
emerging health trends. For example, respondents are asked about their diet and 
weekly physical activity, their HIV/AIDS status, possible tobacco use, and even
their level of healthcare coverage. The BRFSS Web site 
([http://www.cdc.gov/brfss](http://www.cdc.gov/brfss)) contains a complete 
description of the survey, including the research questions that motivate the 
study and many interesting results derived from the data.

We will focus on a random sample of 20,000 people from the BRFSS survey 
conducted in 2000. While there are over 200  variables in this data set, we will
work with a small subset.

We begin by loading the data set of 20,000 observations into the R workspace. 
After launching RStudio, enter the following command.

```{r load-data}
source("http://www.openintro.org/stat/data/cdc.R")
```

The data set cdc that shows up in your workspace (the Environment tab on the top right) is a data matrix, with each row representing a case (one respondent) and each column representing a variable. R calls this data format a data frame, which is a term that will be used throughout the labs.

## Exercise 1

How many cases are there in this data set? How many variables? For each variable, identify its data type (e.g. categorical, discrete).

```{r exercise01}

```

## Exercise 2

Create a numerical summary for height and age, and compute the interquartile range for each. Compute the relative frequency distribution for gender and genhlth. How many males are in the sample? What proportion of the sample reports being in excellent health?

```{r exercise02}

```

## Exercise 3

Create a new object called under23_and_smoke that contains all observations of respondents under the age of 23 that have smoked 100 cigarettes in their lifetime. Write the command you used to create the new object as the answer to this exercise.

```{r exercise03}

```

## Exercise 4

Find the formula for BMI and create a new variable called BMI, corresponding to each person's BMI.

```{r exercise04}

```

## Exercise 5

Make a scatterplot of weight versus desired weight. Describe the relationship between these two variables. 

```{r exercise05}

```

## Exercise 6
Let’s consider a new variable: the difference between desired weight (wtdesire) and current weight (weight). Create this new variable by subtracting the two columns in the data frame and assigning them to a new column called wdiff.

```{r exercise06}

```

## Exercise 7

What type of data is wdiff? If an observation wdiff is 0, what does this mean about the person’s weight and desired weight. What if wdiff is positive or negative?

## Exercise 8

Describe the distribution of wdiff in terms of its center, shape, and spread, including any plots you use. What does this tell us about how people feel about their current weight?

```{r exercise08}

```

## Exercise 9

Using numerical summaries and a side-by-side box plot, determine if men tend to view their weight differently than women.

```{r exercise09}

```

