---
title: "Extra Exercises - for stretching yourself!"
author: "Insert your name here"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## dplyr and ggplot2 with Gapminder

If you'd like to flex your Tidyverse muscles try these exercises that utilise dplyr and ggplot2 with the Gapminder dataset.

```{r load-data}
# Load the tidyverse packages
library(tidyverse)
# gapminder is a subset of data provided by https://www.gapminder.org/ which aims to fight global misconceptions.
library(gapminder)
```

### Exercise 1: Summarizing Data with dplyr

1.  Calculate Average Life Expectancy:

-   Using the gapminder dataset, calculate the average life expectancy (lifeExp) for each continent in the most recent year available.

-   Display the results in descending order of life expectancy.

```{r exercise-1-1}

```

2.  Top 5 Countries by Population:

-   Find the top 5 countries with the largest population (pop) in the most recent year available.

-   Display their names and population.

```{r exercise-1-2}

```

### Exercise 2: Exploring Data with ggplot2

1.  Plot GDP per Capita vs. Life Expectancy:

-   Create a scatter plot of GDP per capita (gdpPercap) vs. life expectancy (lifeExp) for the year 2007.

-   Color the points by continent.

```{r exercise-2-1}

```

2.  Life Expectancy Over Time:

-   Create a line plot showing the trend of life expectancy over time for each continent.
-   Use different colors for each continent and include a smooth trend line.

```{r exercise-2-2}

```

### Exercise 3: Advanced Data Manipulation and Plotting

1.  Continent-wise GDP Growth Rate:

-   Calculate the GDP growth rate from 1952 to 2007 for each continent.
-   Plot the growth rates as a bar plot.

```{r exercise-3-1}

```

2.  Gapminder Data Faceting:

-   Create a faceted scatter plot showing the relationship between GDP per capita and life expectancy for each continent in the year 2007.

-   Each continent should have its own panel.

```{r exercise-3-2}

```

### Further

Explore other datasets that are built into Rstudio like `mtcars` or `iris` using similar techniques.
