---
title: "Week02_TBI"
author: "Kasia Banas"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

You will need to load the required libraries first: tidyverse and here.
Install `here` if you haven't done that already.

```{r libraries}
# install.packages("here")
library(tidyverse)
library(here)
```

We will be using traumatic brain injury data from a Tidy Tuesday challenge
Learn more about the challenge and the data here (spend max 3 minutes looking at the linked page):
https://github.com/rfordatascience/tidytuesday/blob/master/data/2020/2020-03-24/readme.md

Let's start by loading in the data

```{r loading-data}
tbi_age <- read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-03-24/tbi_age.csv')
tbi_year <- read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-03-24/tbi_year.csv')
```

Now, set up folders in your working directory: raw_data, scripts, plots and tables. Put this RMarkdown document in your scripts folder. 

### Task 1
Save both datasets in the raw_data folder (use the 'here' library)

```{r saving-data-locally}

```


### Task 2
Use your favourite functions for eyeballing data and write a couple of sentences about each dataset (its dimensions, the variables inside, types of data, etc.)

### Task 3
Imagine that you are preparing a brief for the local health authority, and they have asked you to answer the following questions. For each question, write a chunk of code that would provide the answer in a clean and reproducible way.

### Task 3.1
Which injury mechanism caused the most deaths in total?

```{r most-deaths}

```

### Task 3.2
Looking just at the totals, what is the number of ED visits, hospitalisations and deaths
associated with each injury mechanism? Present the results as a table. Save this table in your tables folder.

```{r totals-table}
  
```
### Task 3.3
Focusing on the oldest age group (75+), draw a bar chart illustrating the number of hospitalisations associated with each injury mechanism. Make sure that you add a title and appropriate axis labels. Save this chart in your plots folder.

```{r oldest-hospitalisations}

```


