---
title: "Tunour Classifier - K-Nearest Neighbours"
author: "Your name"
date: "September 11, 2024"
output: 
  html_document:
    toc: true
    number_sections: true
editor_options: 
  chunk_output_type: inline
---

## Step 1: Install and load necessary libraries

```{r }
# Install packages if you haven't already
# install.packages("caret")

library(tidyverse)
library(here)
library(caret)
```

## Step 2: Load the dataset

```{r load-data}

cancer_data <- read_csv(here("data", "unscaled_wdbc.csv"))
cancer_data
```

## Step 3: Prepare the data

Encode the target variable, and split the data into training and testing sets.

```{r}

# Keep only the 'Perimeter', 'Concavity', and 'Class' columns
data_subset <- cancer_data %>% 
  select(Perimeter, Concavity, Class)

# Encode Class as a factor
data_subset <- data_subset %>% 
  mutate(Class = factor(Class))
  
# Set a seed for reproducibility
set.seed(42)

# Split the data into training (80%) and testing (20%) sets
train_index <- createDataPartition(data_subset$Class, p = 0.8, list = FALSE)

train_data <- data_subset %>% 
  slice(train_index)

test_data <- data_subset %>% 
  slice(-train_index)

```

## Step 4: Scale the data and train the KNN model

Now, you can train the KNN model using train() from the caret package. You can also use trainControl() to specify cross-validation and scale and center the data if necessary.

```{r}
# Define training control (using 10-fold cross-validation)
trainControl <- trainControl(method = "cv", number = 10)

# Train the KNN model (you can tune 'k' by setting tuneLength)
knn_model <- train(Class ~ ., 
                   data = train_data, 
                   method = "knn",
                   preProcess = c("center", "scale"), 
                   trControl = trainControl, 
                   tuneLength = 10)

# View the results
knn_model

```

## Step 5: Make predictions

Once the model is trained, you can use it to predict on the test set and evaluate its performance.

```{r}
# Make predictions on the test set
predictions <- predict(knn_model, newdata = test_data)

# View the confusion matrix to assess the accuracy
confusionMatrix(predictions, test_data$Class)
```


```{r}
# What if we have a new observation?

new_observation <- tibble(
  Perimeter = 125,
  Concavity = 0.04
) 

prediction <- predict(knn_model, newdata = new_observation)
prediction
```
