---
title: "Stroke mortality table"
author: "Data Science for Health and Biomedical Sciences - Week 6"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```
## Introduction

We will use the package `gt` to create a table with data about stroke mortality in Scotland. If you haven't installed this package yet, please do it now.

We'll be working with a stroke mortality dataset from Public Health Scotland. Follow the link below to find the data dictionary, and a downloadable csv file. https://www.opendata.nhs.scot/dataset/scottish-stroke-statistics/resource/19c01b59-6cf7-42a9-876a-b07b9b92d6eb

You will also need a dataset with health board names, you can find it here: https://www.opendata.nhs.scot/dataset/geography-codes-and-labels/resource/652ff726-e676-4a20-abda-435b98dd7bdc

## Create the table  


Load the libraries
```{r, libraries}
library(tidyverse)  
library(gt) 
library(glue) # This package can be used for the subtitle of the table
```

Import the data
```{r, import data}
 
# Read in the data

stroke_mortality_raw <- read_csv("https://www.opendata.nhs.scot/dataset/f5dcf382-e6ca-49f6-b807-4f9cc29555bc/resource/19c01b59-6cf7-42a9-876a-b07b9b92d6eb/download/stroke_mortalitybyhbr.csv")
 
hb <- read_csv("https://www.opendata.nhs.scot/dataset/9f942fdb-e59e-44f5-b534-d6e17229cc7b/resource/652ff726-e676-4a20-abda-435b98dd7bdc/download/hb14_hb19.csv")
 
```

Join the datasets
```{r, join datasets}
stroke_mortality <- stroke_mortality_raw %>%
  
  # Use left join to join both datasets by health board
  left_join(hb, by = c("HBR" = "HB")) %>%
  
  # Select the variables we are interested in 
  select(Year, HBName, AgeGroup, Sex, Diagnosis, NumberOfDeaths, CrudeRate, EASR) %>% 
  
  # Filter out aggregate levels of the variables 
  filter(Sex != "All" & AgeGroup != "All")
```

Create the table
```{r}

# For the subtitle of the graph, extract the health board names
hb_names <- stroke_mortality %>%
   filter(HBName %in% c("NHS Lothian", "NHS Greater Glasgow and Clyde")) %>%
   distinct(HBName) %>% 
   pull(HBName)

# Create the first version of the table
# Think about filtering, selecting and grouping
stroke_table <- stroke_mortality %>%
  
  # Filter the year, age group, and health board names of interest
  filter(Year == 2018,
          AgeGroup == "75plus years",
          HBName %in% c("NHS Lothian", "NHS Greater Glasgow and Clyde")) %>%
  
  # Group by health board name, and select only the needed columns
  group_by(HBName) %>% 
  select(HBName, 
          Sex, 
          Diagnosis, 
          NumberOfDeaths, 
          CrudeRate, 
          EASR) %>% 
  
  # Generate the table using the gt package
  gt() %>% 
  
  # Move the column 'Sex' after the column 'Diagnosis' 
  cols_move(columns = Sex, after = Diagnosis) %>%
  
  # Rename the columns to make them more user friendly
  cols_label(HBName = "Health Board",
            NumberOfDeaths = "Number of Deaths",
            CrudeRate = "Crude Rate", 
            EASR = "European Age-Sex Standardised Rate") %>% 
  
  # Format the columns with numbers to have two decimal places
  fmt_number(columns = c(NumberOfDeaths, CrudeRate, EASR), decimals = 2) %>% 
  
  # Centre the number columns. Note that the British spelling will not work. Use the
  # American center instead.
  cols_align(align = "center",
             columns = c(NumberOfDeaths, CrudeRate, EASR)) %>% 
  
  # Summarise the number columns to display their means
  summary_rows(columns = c(NumberOfDeaths, CrudeRate, EASR), 
               fns = list("Average" = ~mean(., na.rm = TRUE)),
               fmt = list(~ fmt_number(., decimals = 2))) %>% 
  
  # Add an overall summary of the number columns
  grand_summary_rows(columns = c(NumberOfDeaths, CrudeRate, EASR),
                     fns = list("Overall Average" = ~mean(., na.rm = TRUE)),
                     fmt = list(~ fmt_number(., decimals = 2))) %>% 
  
  # Add a title and subtitle
   tab_header(
       title = "Stroke Mortality in 2018 in Adults Aged 75+",
       
  # For the subtitle we can use the package glue, and paste the unique values of health board
       subtitle = glue("Data from {paste(hb_names, collapse = ' and ')}")) %>%
  
  # Finally, use tab_spanner to add a title to the columns 'CrudeRate' and 'EASR'.
  tab_spanner(
       label = "Rate per 100k population",
       columns = c(CrudeRate, EASR)
   ) 


```



